---
title: "Project 1"
author: "Maxim Averbukh"
date: "11/7/2020"
output: ioslides_presentation
---
<style type="text/css">body p, div, h1, h2, h3, h4, h5 {color:black;font-family:Modern Computer Roman;}slides>slide.title-slide hgroup h1 {color:#8C1D40; <!--the maroon color-->}h2 {color:#8C1D40; <!-- the maroon color-->}</style> <!--end of defining font in various parts of slides-->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What's the Problem?

Samsung carried out an experiment with 30 volunteers. Participants were wearing a Samsung phone which recorder gyroscope and accelerometer data while participants completed 6 basic actions: Walking, walking upstairs, walking downstairs, standing, sitting and lying. 

I would like to use hierarchical clustering to accurately identify what the action completed was based on the data.

The complete data set and more information about it can be found at the link below.  
http://archive.ics.uci.edu/ml/datasets/Smartphone-Based+Recognition+of+Human+Activities+and+Postural+Transitions


## Set up data to work with

First, the data should be transformed to make activity a factor variable
```{r transform, echo = TRUE}
load("samsungData.rda")
samsungData <-transform(samsungData, activity =factor(activity))
```
Now, we can create a subset of data with only the first subject, to make it easier to work with. 
```{r subset, echo = TRUE}
sub1 <-subset(samsungData, subject==1)
```

This data has hundreds of features, so to start I will try to use average acceleration as a way to cluster the data

```{r average cluster, echo = TRUE}
distancebyaverage <-dist(sub1[,1:3])
hclustering <-hclust(distancebyaverage)
```
## Graphing the cluster data

To graph cluster data, I will use a a modified version of the hclust() function from John Hopkins University class "Exploratory Data Analysis".

```{r myplclust, echo = TRUE}
myplclust <-function(hclust, lab = hclust$labels,lab.col
                     =rep(1,length(hclust$labels)), hang = 0.1, ...) 
 {y <-rep(hclust$height, 2)
  x <-as.numeric(hclust$merge)
  y <- y[which(x<0)]
  x <- x[which(x<0)]
  x <-abs(x)
  y <- y[order(x)]
  x <- x[order(x)]
  plot(hclust, labels = FALSE, hang = hang, ...)
  text(x = x, y = y[hclust$order]-(max(hclust$height)*hang), 
       labels = lab[hclust$order],col = lab.col[hclust$order], srt = 90, 
       adj =c(1, 0.5), xpd = NA, ...)}
```
## Graphing the data {#slideID1 .smaller} 
<style> 
  #slideID1 > p { 
    margin-top: -50px; 
  } 
</style>
```{r average graph, echo=FALSE}
myplclust(hclustering, lab.col =unclass(sub1$activity))
```
&nbsp;  
Colors are very mixed so this approach doesn't work as well. Possibly to improve results we could use max acceleration instead of average.

## Graphing the cluster based on max accelration {#slideID2 .smaller} 
<style> 
  #slideID2 > p { 
    margin-top: -50px; 
  } 
</style>
```{r max graph, echo=FALSE}
distancebymax <-dist(sub1[,10:12]) 
hclustering <-hclust(distancebymax)
myplclust(hclustering, lab.col =unclass(sub1$activity))
```
&nbsp;  
This cluster looks much better but some of the motile and all non-motile activities are still mixed. Next improvement strategy is to use SVD.

## Singular Value Decomposition

SVD splits the matrix into a product of three matrices, where A=u*d*v  
- A is the original matrix  
- d is a vector containing the singular values of the original matrix  
- u is a matrix whose columns contain the left singular vectors of the original matrix  
- v is a matrix whose columns contain the right singular vectors of the original matrix   

The decomposed output of SVD provides great insight into variance and pattern distribution in the original matrix.


##  Graphing the cluster from SVD data {#slideID} 
<style> 
  #slideID > p { 
    margin-top: -50px; 
  } 
</style>
```{r svd plot,echo= FALSE}
svdnonames <- svd(scale(sub1[, -c(562, 563)]))#delete non-numeric variables
maxContributer <- which.max(svdnonames$v[, 2])#finds which feature correlates with most variations
##names(samsungData[maxContributer]) #shows maximum contributing feature
distancebysvd <-dist(sub1[,c(10:12,maxContributer)]) #finds distance between the max value of acceleration in x,y and z directions for each data point
hclusteringsvd <-hclust(distancebysvd)#create hclust data 
myplclust(hclusteringsvd, lab.col =unclass(sub1$activity))

```  
  &nbsp;  
This cluster provides a great separation of the 3 moving activities, but all non-moving activities are still mixed together. Next we can try to use k-means clustering. 

## K-means Clustering with 6 groups
```{r kmean6, echo=TRUE}
kmean6 <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 100)
table(kmean6$cluster,sub1$activity)
```
In the above table it is clear that the 3 motile activities are well distinguished but the non-motile activities are all mixed together. 

## K-means cluster results compared to the actual data {.smaller #slideID4} 
<style> 
  #slideID4 > p { 
    margin-top: -50px; 
  } 
</style>
K-means table:
```{r kmean table}
table(kmean6$cluster,sub1$activity)
```
  &nbsp;  
Actual data table:
```{r data table}
table(sub1$activity)
```
  &nbsp; 
  &nbsp; 
  &nbsp;  
We see that the motile activities are exactly on point. Let's try graphing this cluster for a better visualization.

## K-means Plot With 6 Groups {.smaller #slideID5} 
<style> 
  #slideID5 > p { 
    margin-top: -50px; 
  } 
</style>
```{r kmean plot, message=FALSE, warning=FALSE}
library("factoextra")
fviz_cluster(kmean6, sub1[, -c(562, 563)])
```
  &nbsp;  
  We can see that the algorithm is still struggling to distinguish the non-motile activities from each other. We can try K-means again but with only 4 groups, so that all non-motile data is just bundled together.
  
  
## K-means Clusterin with 4 Groups{.smaller #slideID6} 
<style> 
  #slideID6 > p { 
    margin-top: -50px; 
  } 
</style>
```{r kmean4 }
kmean4 <- kmeans(sub1[, -c(562, 563)], centers = 4, nstart = 100)
fviz_cluster(kmean4, sub1[, -c(562, 563)])
```
 &nbsp;  
 The data looks much better with only 4 groups, where all non-motile activities are in 1 group.
 
## Dendogram with 4 groups {.smaller #slideID7} 
<style> 
  #slideID7 > p { 
    margin-top: -50px; 
  } 
</style>
 ```{r clust4  }
 clust4 <- hcut(sub1[, -c(562, 563)], k = 4, stand = TRUE)
fviz_dend(clust4, rect = TRUE, cex = 0.5)
```
 &nbsp;  
The dendogram using Hierarchical Multiple Factor Analysis also shows a very clear visualization of the data. As a future direction for this project, we could try and find other machine learning approaches to classify the non-moving data as well.
